Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/VIRGO Network

### 2013-05-14 doc/paper.tex
rough draft

%\IfFileExists{revtex4-1.cls}{\documentclass[prd, twocolumn, lengthcheck, superscriptaddress, showpacs, letterpaper, footinbib]{revtex4-1}}{ 
%\IfFileExists{revtex4.cls}{\documentclass[prd, twocolumn, lengthcheck, superscriptaddress, showpacs, letterpaper]{revtex4}}{}}
\documentclass{emulateapj}

%\documentclass[prd, twocolumn, lengthcheck,superscriptaddress, showpacs, letterpaper, footinbib]{revtex4}
%\documentclass[lengthcheck,superscriptaddress,showpacs,letterpaper,nofootinbib]{article}
\usepackage{ifpdf}
\usepackage[bookmarks=false]{hyperref}
\pdfoutput=1
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx}
\usepackage{graphics}  
  \usepackage{float}
%\restylefloat{table}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{acronym}
\usepackage{multirow}
\usepackage{xspace}
%\restylefloat{figure}
\usepackage{units}
\usepackage{dcolumn}
\usepackage{ulem}

@@

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/VIRGO Network}
\author{Carl L. Rodriguez}
\email{carlrodriguez2015@u.northwestern.edu}
\affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \& Dept.~of Physics and Astronomy, Northwestern University, 2145 Sheridan Rd, Evanston, IL 60208, USA}

\begin{abstract}
\end{abstract}

@@

 By 2016 (\carl{?}), the field of gravitational-wave astrophysics will mature from a speculative scientific genre into an ideal tool for exploring the gravitational side of the universe.
 Within the next few years, the first generation of gravitational-wave detectors capable of regularly resolving astrophysical sources will come online \carl{cite}.   The Advanced LIGO and VIRGO detectors will provide  the first insights into the final moment of binary compact object mergers, including the merger of binary neutron star systems.  Intense preparations are currently underway to prepare to characterize and extract as much physical information as possible from these signals.

The mergers of binary neutron stars are expected to be the most ubiquitous source in the advanced detector era \carl{cite}.   Models from stellar evolution suggest that the number of binary neutron star mergers within Advanced LIGO/VIRGO's detection horizon could reach into the hundreds per year \carl{cite}.  Although the peak sensitivity of ground-based detectors is not focused on the frequency at which BNS systems merge, it should still be possible to extract information about both strong field gravitational physics \carl{cite} and the hydrodynamics of dense matter (e.g. the equation of state of nuclear matter) \carl{cite}.  As such, BNS systems will likely form the ``bread and butter'' of the compact object efforts in the coming years \carl{the high-mass folks might not like this, but fuck 'em}.

Of course, in this context, one must distinguish between the detection of such events, and the precision measurement of the relevant physical parameters.   There are two separate issues to consider in this context: the detection of BNS systems will be performed with a matched filtering approach, which allows one to search the time domain data at a sufficiently fast rate \carl{[yeah that phrasing sucks]} by comparing the data stream to a bank of theoretical templates.  When the overlap between one of the template waveforms and the detector output is sufficiently large (97\%), a detection is claimed.  However, the parameter space of our signals can be highly degenerate, with several locations in parameter space corresponding to near identical waveforms.  In order to fully realize the science potential of these instruments, we must perform a full exploration of the parameter space for each detection.  By analyzing each candidate's parameter space with fully Bayesian techniques and proper sampling, we will be able to make precise, scientifically meaningful statements about the composition and formation of these objections.  To this end, we employ a Markov-Chain Monte Carlo sampling algorithm, included in the LIGO Application Library parameter estimation code, \texttt{LALInferenceMCMC}.

By employing the full parameter estimation machinery that will eventually be used in the Advanced Detector era, our results give the first estimate at the realistic capabilities of Advanced ground-based detectors.  Until recently, the majority of studies have employed the Fisher matrix formalism as it was first adapted for gravitational-wave parameter estimation by \cite{FinnDetection}.  While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among many others) have done their due diligence in pointing out the limitations and flaws of the Fisher Information Matrix, there has been no study which states the BNS parameter estimation capabilities of Advanced LIGO/VIRGO via the techniques that will eventually be employed.  The recent work of \cite{Vallisneri} and our own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even be treated as a lower bound on the standard deviations about certain parameters (particularly the masses) measurable in advanced detectors.

In this paper, we perform a systematic study of the statistical uncertainty with which the Advanced LIGO network will be able to measure the basic parameters of binary neutron stars.  In section \ref{BNSsection}, we review some of the basic physics of gravitational waves from binary neutron stars, including the potential science case to be made with Advanced LIGO.  In section \ref{PEsection}, we describe the machinery of our parameter estimation code, \texttt{LALInferenceMCMC}, as well as the gravitational-wave template used for signal extraction.   In section \ref{ResultsSection}, we

\begin{itemize}
 \item BNS source
\item PE vs detection
\item Why not FIM
\item Overview
\end{itemize}

\section{BNS systems}
\label{BNSsection}
\begin{itemize}
\item expected rates
\item interesting physics
\item Expected parameters (non spinning, SNR distribution, etc)
\item waveform issues
\end{itemize}

@@

\begin{itemize}
\item parameter estimation setup
\item MCMC
\item Waveform model
\end{itemize}
\label{peSection}

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in our gravitational wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with mean zero.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise is assumed to be Gaussian, we can write the
probability of a specific signal realization $s$ as the probability
that our signal is Gaussian once the waveform has been subtracted
\begin{align}
  p(s | \boldsymbol{\theta}) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta})\right >\right] ,
  \label{likelihood}
\end{align}
where $\boldsymbol{\theta}$ is the set of parameters for our template
waveforms.The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Rule to obtain the posterior probability of the system
parameters $\boldsymbol{\theta}$ given the signal $s$ as
\begin{align}
  p(\boldsymbol{\theta} | s) &= \frac{p(\boldsymbol{\theta})p(s | \boldsymbol{\theta})}{p(s)} \nonumber\\
  & \propto p(\boldsymbol{\theta}) \exp\left[-\frac{1}{2}\big < s - h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta}) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\boldsymbol{\theta})$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.  Our prior
information can come from physics limits in the parameter space, or
from our \textit{a priori} knowledge of astrophysical systems. If we
pick a set of parameters $\boldsymbol{\theta}$ such that
$h(\boldsymbol{\theta}) = h_0$, then our posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\boldsymbol{\theta}_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.  Hence our use of the term
``best-fit parameters'' (\citep{CutlerFlanagan}), particularly when
discussing the point in parameter space where the FIM is evaluated.
The goal of parameter estimation is to sample the available parameter
space, considering all areas of posterior support, to determine best
estimates on the parameters of the signal and their uncertainties.  We
will discuss one such sampling technique, Markov-Chain Monte Carlo, in
\ref{MCMCSection}.

\subsection{Parameter estimation via Markov-Chain Monte Carlo}
\label{MCMCSection}

Our parameter estimation code, \texttt{LALInferenceMCMC}, is an
enhancement of the previously described MCMC parameter estimation code
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The basic
description of the Markov-Chain Monte Carlo via the
Metropolis-Hastings algorithm is as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\boldsymbol{\theta}|s\right)$.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary phase approximation \citep{BuonannoWaveform}.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = A f^{-7/6}e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $A \propto \mathcal{M}_c^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M\equiv m_1+m_2$, it is convenient to work with the
\textit{symmetric mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  \eta\equiv m_1m_2/M^2~~~~\text{and}~~~~\mathcal{M}_c = \eta^{3/5}M,
\end{equation}
respectively.  Then, in terms of the Newtonian orbital velocity
$v=(\pi M f)^{1/3}$, the 2pN phase is
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128 \eta}v^{-5}\sum^{4}_{k=0}\alpha_{k}v^k
\label{phase}
\end{equation}
with coefficients
\begin{align}
\alpha_0 &= 1\\
\alpha_1 &= 0 \nonumber\\
\alpha_2 &= \frac{20}{9}\left(\frac{743}{336}+\frac{11}{4}\eta\right)\nonumber\\
\alpha_3 &= -16\pi \nonumber\\
\alpha_4 &= 10\left(\frac{3058673}{1016064}+\frac{5249}{1008}\eta+\frac{617}{144}\eta^2  \right) . \nonumber\\
\alpha_5 &= \pi\left(\frac{38645}{756}-\frac{65}{9}\eta\right)\left(1+3\log\left(\frac{v}{v_{\text{ISCO}}}\right)\right)\nonumber\\
\alpha_6 &= \frac{11583231236531}{4694215680} - \frac{640}{3}\pi^2 - \frac{6848}{21}\gamma_{\text{euler}} \nonumber\\
         &~~- \frac{6848}{21} \log\left(4 \frac{v}{v_{\text{ISCO}}}\right) -\left(\frac{15737765635}{3048192}-\frac{2255}{12}\pi^2\right)\eta\nonumber\\
           &~~+ \frac{76055}{1728}\eta^2 - \frac{127825}{1296}\eta^3\nonumber\\
\alpha_7 &= \pi\left(\frac{77096675}{254016}+\frac{1014115}{3024}\eta - \frac{36865}{368}\eta^2\right)\nonumber
\end{align}
The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants
of integration, referring to the chirp time and coalescence phase,
respectively.  Although uninteresting physically, they must be
accounted for in any parameter estimation study of the waveform phase.

To perform the integral defined in \eqref{innerProduct}, we used an
analytic noise curve for the Initial LIGO configuration, provided in
\citet{DarmourNoiseCurve}, which takes the form:
\begin{align}
 S_h(f)=&S_0\Big[ \left(\frac{4.49 f}{f_0} \right)^{-56} + \nonumber\\
 &\left(\frac{0.16f}{f_0}\right) ^{-4.52}+0.52+0.32\left( \frac{f}{f_0}\right) ^2\Big],
 \label{PSD}
\end{align}
where $f_0 = 150~\text{Hz}$, and $S_0 = 9\times
10^{-46}\text{Hz}^{-1}$.  We consider the complete initial detector
network consisting of the two LIGO sites (in Hanford, WA and
Livingston, LA) and the Virgo site (in Pisa), although for simplicity
we use the Initial LIGO sensitivity for all three detectors.  For a
multi-detector network, the noise-weighted inner products
\eqref{innerProduct} combine linearly, allowing us to use the above
formalism with minimal modification.  We integrate the inner product
from a lower frequency cutoff of $40\text{Hz}$ to the
innermost-stable-circular orbit of the system in question, which for a
non-spinning binary is given by
\begin{equation}
  \pi f_{\text{ISCO}} = \frac{1}{6^{3/2}M}.
  \label{ISCOFrequency}
\end{equation}

In addition to the intrinsic parameters listed above ($\mathcal{M}_c,
\eta, \phi_0,t_c$), we include an additional 5 extrinsic parameters in
our parameter estimation, in order to explore how a full parameter
estimation study on real data, including parameters such as sky
location, will compare to their FIM counterparts.  This leads to our
9-dimensional parameter space for non-spinning systems:
\begin{equation}
\boldsymbol{\theta} =  (\mathcal{M}_c, \eta, \phi_0,t_c,D,\iota,\psi,\alpha,\delta)
\label{parameterspace}
\end{equation}
where $D$ is the luminosity distance to the binary, $\iota$ is the
orbital inclination, $\psi$ the gravitational-wave polarization, and
$\alpha$ and $\delta$ the right ascension and declination of the
source on the sky.

Note that the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector is
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}^{*}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively.  The SNRs for
multiple detectors add in quadrature.  The normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}
For our MCMC analysis, we compute the SNR directly using
\eqref{formalSNR}.  In the case of the FIM, we approximate the SNR
using $\sigma$ from equation (\ref{SNRnorm}) for a network of
detectors as
\begin{equation}
\rho = \sqrt{\sum_i \sigma_i^2}
\label{snr}
\end{equation}
where the index $i$ refers to the signal and noise spectrum of the
$i^{\text{th}}$ detector.  It should be noted that equation
\eqref{snr} is only valid in case of stationary Gaussian noise.

>>>>

%\IfFileExists{revtex4-1.cls}{\documentclass[prd, twocolumn, lengthcheck, superscriptaddress, showpacs, letterpaper, footinbib]{revtex4-1}}{ 
%\IfFileExists{revtex4.cls}{\documentclass[prd, twocolumn, lengthcheck, superscriptaddress, showpacs, letterpaper]{revtex4}}{}}
\documentclass[11pt,a4paper]{emulateapj}
\bibliographystyle{apj}

\usepackage{epsfig}
%\usepackage{amsmath}
\usepackage{natbib}

%\documentclass[prd, twocolumn, lengthcheck,superscriptaddress, showpacs, letterpaper, footinbib]{revtex4}
%\documentclass[lengthcheck,superscriptaddress,showpacs,letterpaper,nofootinbib]{article}
\usepackage{ifpdf}
\usepackage[bookmarks=false]{hyperref}
\pdfoutput=1
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx}
\usepackage{graphics}  
  \usepackage{float}
%\restylefloat{table}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{acronym}
\usepackage{comment}
\usepackage{multirow}
\usepackage{xspace}
%\restylefloat{figure}
\usepackage{units}
\usepackage{dcolumn}
\usepackage{ulem}

@@

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/VIRGO Network}
\author{Carl L. Rodriguez \altaffilmark{1}
%\email{rodriguez@u.northwestern.edu}
,Benjamin Farr \altaffilmark{1}
%\email{bfarr@u.northwestern.edu} 
Vivien Raymond \altaffilmark{1,2}
Will M. Farr \altaffilmark{1}
%\email{w-farr@northwestern.edu} 
Tyson Littenberg\altaffilmark{1}
%\email{tyson.littenberg@northwestern.edu}
Diego Fazi\altaffilmark{1}
%\email{d-fazi@northwestern.edu}
Vicky Kalogera\altaffilmark{1}}
%\email{vicky@northwestern.edu}

\altaffiltext{1}{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \& Dept.~of Physics and Astronomy, Northwestern University, 2145 Sheridan Rd, Evanston, IL 60208, USA; [e-mail: {\tt cr@u.northwestern.edu}]}
\altaffiltext{2}{California Institute of Technology, Pasadena, CA 92215, USA}

\begin{abstract}

Within the next 5 years, the Advanced LIGO/VIRGO network will have reached a sensitivity sufficient to enable the routine detection of gravitational waves.  Beyond the initial detection, the scientific promise of these instruments relies on the efficiency of the parameter estimation techniques being currently developed.  The majority of this effort has been towards the detection and characterization of gravitational waves from compact binary coalescence, i.e. the coalescence of black holes and/or neutron stars.  While several previous studies have focused on the parameter estimation abilities of advanced detectors, the vast majority have relied on approximation techniques (such as the Fisher Matrix) when characterizing the uncertainties.  Here, we report the statistical uncertainties that will be achievable, in practice, using the parameter estimation code \texttt{LALInference} to recover the parameters of binary neutron star systems detected in various configurations of the Advanced LIGO/VIRGO network.  We find that the recovery of the individual masses will be, on average, sufficient to differentiate neutron stars from black holes by the masses alone.  We also report the average uncertainties on the sky-locations, luminosity distance, and orbital inclinations of randomly generated sources that can be recovered by the expected three and four-detector networks.
\end{abstract}

@@

 By 2016, the field of gravitational-wave astrophysics will
 mature from a speculative scientific genre into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and VIRGO detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense preparations are currently
 underway to prepare to characterize and extract as much physical 
 information as possible from these signals.

The mergers of binary neutron stars are expected to be the most ubiquitous
source in the advanced detector era \carl{[CITE]}.   Models from stellar 
evolution suggest that the number of binary neutron star mergers within 
Advanced LIGO/VIRGO's detection horizon could reach into the hundreds per 
year \carl{[CITE]}.  Although the peak sensitivity of ground-based detectors
is not focused on the frequency at which BNS systems merge, it should still
be possible to extract information about both strong field gravitational 
physics \carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation 
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the observations of multiple BNS
systems will provide key insight into the evolution of binary systems in the field \carl{[CITE]},
and in globular clusters \carl{[CITE]}. As such, BNS systems will likely 
form the ``bread and butter'' of the compact object efforts in the coming 
years.

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream to a bank of theoretical templates, the roughly 16KHz
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection is claimed.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to near identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
and proper sampling, we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systemss.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/VIRGO era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have done their due diligence in pointing out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which states
the BNS parameter estimation capabilities of Advanced LIGO/VIRGO via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations about certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

In this paper, we perform a systematic study of the statistical uncertainties with
which the Advanced LIGO/VIRGO network will be able to measure the basic parameters of 
binary neutron stars.  In Section \ref{PEsection}, we describe
the machinery of our parameter estimation code, \texttt{LALInference}, and the associated
MCMC sampler, \texttt{lalinference\_mcmc}, as well as the frequency-domain gravitational-wave template we employ.   
In Section \ref{resultsSection}, we qualitatively analyze the posterior probability density functions for BNS systems with different masses and extrinsic parameters.  The results are divided into three parameter sets of interest: the recovery of the mass parameters (Section \ref{massSection}), the recovery of the orbital inclination and luminosity distance (Section \ref{idSection}), and the isolation of sources in the sky (Section \ref{skySection}).   Finally we provide quantitative 1-dimensional confidence intervals on the parameter recovery in Section \ref{ciSection}.  We assume that $G=c=1$ throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in our gravitational wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with mean zero.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as the probability
that our signal is Gaussian once the waveform has been subtracted
\begin{align}
  p(s | \boldsymbol{\theta}) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta})\right >\right] ,
  \label{likelihood}
\end{align}
where $\boldsymbol{\theta}$ is the set of parameters for our template
waveforms.  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Rule to obtain the posterior probability of the system
parameters $\boldsymbol{\theta}$ given the signal $s$ as
\begin{align}
  p(\boldsymbol{\theta} | s) &= \frac{p(\boldsymbol{\theta})p(s | \boldsymbol{\theta})}{p(s)} \nonumber\\
  & \propto p(\boldsymbol{\theta}) \exp\left[-\frac{1}{2}\big < s - h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta}) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\boldsymbol{\theta})$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.  Our prior
information can come from physics limits in the parameter space, or
from our \textit{a priori} knowledge of astrophysical systems. If we
pick a set of parameters $\boldsymbol{\theta}$ such that
$h(\boldsymbol{\theta}) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\boldsymbol{\theta}_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

\subsection{Markov-Chain Monte Carlo}
\label{MCMCSection}

The LIGO Algorithm Library parameter estimation code, \texttt{LALInference}, is designed as 
a unified framework for gravitational-wave parameter estimation.  By using a common setup for 
waveform generation, PSD estimation, data handling, and other associated techniques from gravitational-wave parameter estimation, \texttt{LALInference} allows the implementation of multiple parameter samplers in using identical parameter setups, including nested sampling (\texttt{lalinference\_nest}) and Markov-Chain Monte Carlo (\texttt{lalinference\_mcmc}).  We elect to use the MCMC sampler for this study.  \texttt{lalinference\_mcmc} is based upon the previously described code, 
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The MCMC employs a Metropolis-Hastings sampling algorithm, which can be described as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\boldsymbol{\theta}|s\right)$.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/VIRGO network.  
A description of the parameter estimation capabilities, c.f. 2010, can be found
in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = A f^{-7/6}e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $A \propto \mathcal{M}_c^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\mathcal{M}_c = (M_1 M_2)^{3/5} M^{-1/5},
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead use the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is empirically superior when proposing MCMC parameter jumps, and is more physically intuitive.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\mathcal{M}_c}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

\noindent where the $\alpha_{k}$ coefficients are taken from the pN expansion to order $n/2$.  See \cite{BuonannoWaveform} for a description and comparison of different waveform families.

The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants
of integration, referring to the chirp time and coalescence phase,
respectively.  Although uninteresting physically, they must be
accounted for in any parameter estimation study of the waveform phase.

To perform the integral defined in \eqref{innerProduct}, we used the best estimate
for a high-power, zero-detuning configuration of Advanced LIGO, provided by the LIGO Scientific Collaboration.
Both the noise curve and technical reports describing it can be found in \cite{ADVLIGONoise}.
We consider two configurations of the advanced detector
network: a three-detector configuation consistanting of the two LIGO sites 
(in Hanford, WA and Livingston, LA) and the Virgo site (in Pisa), and a four-detector
configuration with adds the proposed LIGO-India detector (in Chitradurga, KA).
For simplicity, we assume each detector to be operating at the Advanced LIGO sensitivity. 
For a multi-detector network, the noise-weighted inner products
\eqref{innerProduct} combine linearly, allowing us to use the above
formalism with minimal modification.  We integrate the inner product
from a lower-frequency cutoff of $20\text{Hz}$ to the
innermost-stable-circular orbit of the systems in question, which for a
non-spinning binary is a function only of the total mass:

\begin{equation}
  \pi f_{\text{ISCO}} = \frac{1}{6^{3/2}M}.
  \label{ISCOFrequency}
\end{equation}

We consier the mass parameters ($\mathcal{M}_c,q$) and the phasing parameters,
($\phi_0,t_c$), to be the intrinsic parameters of a gravitational-wave signal.  There 
are an additional 5 extrinsic parameters, independent of the pN orbital phase, which can
modulate the amplitude of the signal in an independent fashion for each detector.  
Combining these leads to our 9-dimensional parameter space for non-spinning systems:
\begin{equation}
\boldsymbol{\theta} =  (\mathcal{M}_c, q, \phi_0,t_c,D,\iota,\psi,\alpha,\delta)
\label{parameterspace}
\end{equation}
where to summarize

\begin{itemize}
\item $\mathcal{M}_c$ is the chirp mass,
\item $q$ is the mass ratio,
\item $\phi_0$ and $t_c$ are the chirp phase and chirp time, arbitrary phasing parameters,
\item $D$ is the luminosity distance to the binary,
\item $\iota$ is the orbital inclination, 
\item $\psi$ the gravitational-wave polarization, and
\item $\alpha$ and $\delta$ the right ascension and declination of the
source on the sky.
\end{itemize}
  Since the wave amplitude depends on the orientation of the
binary with respect to each detector, most of the information about these extrinsic
parameters comes from two sources: the time-of-arrival triangulation of 
the source locations, and the relative amplitudes within the detector network.

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector is
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}^{*}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

When dealing with network of gravitational-wave detectors, the SNRs of the individual detectors add in quadrature.  That is, the network SNR for a detection is

\begin{equation}
\rho_{\text{network}} = \sqrt{\sum_i \rho_{i}}
\label{SNRnetwork}
\end{equation}

\noindent where $\rho_i$ is the SNR, given by \eqref{formalSNR}, of the $i^{\text{th}}$ detector.

For this study, we consider four seperate populations of BNS systems, with component mass 
combinations of $1M_{\odot}/1M_{\odot}$, $1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, 
and $2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals distributed randomly
in sky location, polarization, inclination, time-of-arrival, and coalescence phase.  The luminosity 
distance, $D$, was selected to provide a network SNR of $\rho_{\text{network}}=20$ for each source.
While unphysical from an astrophysical perspective (binary sources should be 
distributed evenly in comoving volume), this choice allows us to explore the parameter estimation
in the context of viable detection candidates.

Finally, note that the power-spectral density that defines the inner product \eqref{innerProduct} is simply 
the time-averaged sensitivity of a given detector to a specific frequency.  Ignoring non-Gaussian glitches,
 any stretch of data should contain a specific noise realization drawn from a Gaussian colored by the PSD.
   However, what we are interested in is not the parameter estimation that can occur in a specific draw
    from this probability of noises, but the uncertainties averaged over all possible noise realizations.
    \footnote{Similarly to how the Fisher matrix formalism can provide the Cramer-Rao bound averaged over
     all noise realizations} Since our noise model is assumed to be Gaussian, it is unnecessary to recover the
      parameters from a large sample of identical waveforms in different noise; we can simply assume the 
      noise \emph{is} the mean defined by the power-spectral density.  This ``zero-noise'' parameter estimation
       provides a statical statement on the measurable uncertainties, and is what is reported here.  In Section
        \ref{noiseSection} we provide an example of parameter estimation in several non-zero noise realizations
compared to the zero-noise equivelant.
